---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# AssessR Package

An R package to measure the performance of property assessment using standard statistics.
Also includes a host of utility functions to make common assessment and evaluation tasks easier and more consistent.

## Installation

You can install the released version of assessR directly from GitLab by running the following R command after installing the `remotes` package:

```{r, eval=FALSE}
remotes::install_git("https://gitlab.com/ccao-data-science---modeling/packages/assessr")
```

Once it is installed, you can use it just like any other package. Simply
call `library(assessr)` at the beginning of your script.

## Example Calculations

Using the included `ratios_sample` dataset, `cod()`, `prd()`, and `prb()` can be used to measure the performance of an assessment.

```{r, message=FALSE}
library(dplyr)
library(assessr)
library(knitr)

# Load the sample dataset
data("ratios_sample")

# Calculate peformance statistics by township
ratios_sample %>%
  group_by(town) %>%
  summarize(
    cod = cod(ratio),
    prd = prd(assessed, sale_price),
    prb = prb(assessed, sale_price)
  ) %>%
  kable(format = "markdown")

```

To add:
- Coefficient of variation
- Outlier detection
- Bootstrap confidence intervals
- sales chasing detection
- trimming functions
- sample size verification
- sales chasing prevention
