---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# AssessR Package

An R package to measure the performance of property assessment using standard statistics.
Also includes a host of utility functions to make common assessment and evaluation tasks easier and more consistent.

## Installation

You can install the released version of assessR directly from GitLab by running the following R command after installing the `remotes` package:

```{r, eval=FALSE}
remotes::install_git("https://gitlab.com/ccao-data-science---modeling/packages/assessr")
```

Once it is installed, you can use it just like any other package. Simply
call `library(assessr)` at the beginning of your script.

## Example Calculations

Using the included `ratios_sample` dataset, `cod()`, `prd()`, and `prb()` can be used to measure the performance of an assessment.

```{r, message=FALSE, results='asis'}
library(dplyr)
library(assessr)
library(knitr)

# Load the sample dataset
data("ratios_sample")

# Calculate peformance statistics by township
ratios_sample %>%
  group_by(town) %>%
  summarize(
    cod = cod(ratio),
    cod_ci = paste(round(cod_ci(ratio, nboot = 1000), 3), collapse = ", "),
    prd = prd(assessed, sale_price),
    prd_ci = paste(round(prd_ci(assessed, sale_price), 3), collapse = ", "),
    prb = prb(assessed, sale_price),
    prb_ci = paste(round(prb_ci(assessed, sale_price), 3), collapse = ", ")
  ) %>%
  rename_all(toupper) %>%
  kable(format = "markdown", digits = 3)

```

## Using Real Data

This package can easily be used with data from the [Cook County Open Data Portal](https://datacatalog.cookcountyil.gov/Property-Taxation/Cook-County-Assessor-s-Residential-Assessments/uqb9-r7vn) to analyze assessment performance. To measure assessment performance, you will need to gather both sales and assessed values. These are stored in two separate datasets on the data portal.

### With RSocrata

[RSocrata](https://github.com/Chicago/RSocrata) is a package developed by the City of Chicago to wrap Socrata API requests. It allows you to easily pass a Socrata app token, which will remove the API limit on the number of rows returned. Example usage is shown below, replacing the login details with your own.

```{r, eval=FALSE}
library(RSocrata)

# Load unlimited rows of assessment data, default is 1,000
assessments <- read.socrata(
  "https://datacatalog.cookcountyil.gov/resource/uqb9-r7vn.json",
  app_token = "YOURAPPTOKENHERE",
  email     = "user@example.com",
  password  = "fakepassword"
)
```

### With jsonlite

Socrata can also return raw JSON if you manually construct a query URL. Follow the [API docs](https://dev.socrata.com/foundry/datacatalog.cookcountyil.gov/uqb9-r7vn) to alter your query. The raw JSON output can be read using the `read_json()` function from `jsonlite`.

```{r}
library(jsonlite)

# Load 100k rows of assessment data
assessments <- read_json(
  "https://datacatalog.cookcountyil.gov/resource/uqb9-r7vn.json?$limit=100000",
  simplifyVector = TRUE
)

# Load 100k rows of sales data
sales <- read_json(
  "https://datacatalog.cookcountyil.gov/resource/5pge-nu6u.json?$limit=100000",
  simplifyVector = TRUE
)


```

### Example Analysis

Using the collected assessment and sales data, we can perform a rudimentary analysis and measure the performance of each town ship at each stage of assessment. 

```{r, results='asis'}
library(dplyr)
library(tidyr)
library(knitr)

# Join the two datasets based on PIN, keeping only properties that have assessed
# values AND sales
combined <- inner_join(
  assessments %>% select(pin, year, town_name, first_pass, certified, bor_result),
  sales %>% select(pin, year = sale_year, sale_price),
  by = c("pin", "year")
)

# Remove sales that are not arms length, pivot to longer, then calculate 
# the ratio for each property and assessment stage
combined <- combined %>% 
  filter(sale_price >= 10000) %>%
  pivot_longer(
    first_pass:bor_result,
    names_to = "stage",
    values_to = "assessed"
  ) %>%
  mutate_at(vars(sale_price, assessed), as.numeric) %>%
  mutate(ratio = assessed / sale_price)


# For each town and stage, calculate COD, PRD, and PRB, and their respective
# confidence intervals then arrange by town name and stage of assessment
combined %>%
  group_by(town_name, stage) %>%
  summarise(
    n = n(),
    cod = cod(ratio),
    cod_ci = paste(round(cod_ci(ratio, nboot = 1000), 3), collapse = ", "),
    prd = prd(assessed, sale_price),
    prd_ci = paste(round(prd_ci(assessed, sale_price), 3), collapse = ", "),
    prb = prb(assessed, sale_price),
    prb_ci = paste(round(prb_ci(assessed, sale_price), 3), collapse = ", ")
  ) %>%
  filter(n >= 50) %>%
  mutate(stage = factor(
    stage,
    levels = c("first_pass", "certified", "bor_result"))
  ) %>%
  arrange(town_name, stage) %>%
  rename_all(toupper) %>%
  kable(format = "markdown")

```

